\documentclass[a4paper,10pt, notitlepage]{report}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[shortlabels]{enumitem}
% \usepackage[portuguese]{babel}


% Title Page
\title{O Brother, How Far Art Thou?}
\author{Computational Statistics \\ Instructor: Luiz Max de Carvalho}

\begin{document}
\maketitle

\textbf{Hand-in date: 06/10/2021.}

\section*{General guidance}
\begin{itemize}
 \item State and prove all non-trivial mathematical results necessary to substantiate your arguments;
 \item Do not forget to add appropriate scholarly references~\textit{at the end} of the document;
 \item Mathematical expressions also receive punctuation;
 \item Please hand in a single PDF file as your final main document.
 
 Code appendices are welcome,~\textit{in addition} to the main PDF document.
 \end{itemize}

\newpage

\section*{Background}

A large portion of the content of this course is concerned with computing high-dimensional integrals~\textit{via} simulation.
Today you will be introduced to a simple-looking problem with a complicated closed-form solution and one we can approach using simulation.

Suppose you have a disc $C_R$ of radius $R$. 
Take $p = (p_x, p_y)$ and $ q = (q_x, q_y) \in C_R$ two points in the disc.  
Consider the Euclidean distance between $p$  and $q$, $||p-q|| = \sqrt{(p_x-q_x)^2 + (p_y-q_y)^2} = |p-q|$.
\paragraph{Problem A:} What is the \textit{average} distance between pairs of points in $C_R$ if they are picked uniformly at random?

\section*{Part I: nuts and bolts}

\begin{enumerate}
 \item To start building intuition, let's solve a related but much simpler problem.
 Consider an interval $[0, s]$, with $s>0$ and take $x_1,x_2 \in [0, s]$~\textit{uniformly at random}.
 Show that the average distance between $x_1$ and $x_2$ is $s/3$.
 \item Show that Problem A is equivalent to computing
 \begin{equation*}
  I = \frac{1}{\pi^2 R^4}\int_{0}^{R}\int_{0}^{R}\int_{0}^{2\pi}\int_{0}^{2\pi}\sqrt{r_1^2 + r_2^2 - 2r_1r_2\cos\phi(\theta_1, \theta_2)}r_1r_2\,d\theta_1\,d\theta_2\,dr_1\,dr_2,
 \end{equation*}
 where $\phi(\theta_1, \theta_2)$ is the central angle between $r_1$ and $r_2$.
 
 \textit{Hint:} Draw a picture.
 \item Compute $I$ in closed-form.

 \textit{Hint:} Look up \textit{Crofton's mean value theorem} or \textit{Crofton's formula}. 
\end{enumerate}

\section*{Part II -- getting your hands dirty}

Now we will move on to implementation.

\paragraph{Problem B:} Employ a simulation algorithm to approximate $I$.
Provide point and interval estimates and give theoretical guarantees about them (consistency, coverage, etc).

\begin{enumerate}
 \item You have been (randomly) assigned a simulation method -- see list at the end.
 Represent $I$ as $\int_{\mathcal{X}} \phi(x)\pi(x)\,dx$ and justify your choice of $\phi$, $\pi$ and $\mathcal{X}$.
 Recall that these choices are arbitrary up to a point, but they might lead to wildly different empirical performances~\textbf{and} theoretical properties for estimators of $I$.
 \textbf{Justify} your choices in light of the method you have been given to work with.
 Choose wisely and be rigorous in your justifications.
 \item Again, starting from the eventual samples you will obtain with your method, construct a non-empty\footnote{This is a joke.
 It means you should come up with at least one estimator. But you might, and are even encouraged to, entertain more than one estimator.} family of estimators of $I$ and discuss whether it is (strongly) consistent and whether a central limit theorem can be established.
 \item Detail a suite of diagnostics that might be employed in your application to detect convergence or performance problems.
 Extra points for those who design algorithms that exploit the structure of this particular integration problem. 
 \item For each $R \in \{0.01, 0.1, 1, 10, 100, 1000, 10000\}$, perform $M=500$ runs from your simulation method and compute: (i) variance (ii) bias (iii) standard deviation of the mean (MCSE).
 \item Can you identify one key quantity missing from the previous item?
 \textit{Hint:} it bears relevance to the real world application of any computational method.
estimator.\end{enumerate}

Here we will list a selection of methods that will be randomly assigned to each student, along with some questions that need to be answered for that particular method.

\begin{itemize}
 \item \textbf{Rejection sampling}
  \begin{itemize}
  \item Justify your choice of proposal distribution and show that it conforms to the necessary conditions for the algorithm to work; in particular, try to find a proposal that gives the highest acceptance probability.
 \end{itemize}
  \item \textbf{Importance sampling}
 \begin{itemize}
  \item Justify your choice of proposal based on the variance of the resulting estimator.
 \end{itemize}
  \item \textbf{Gibbs sampling}
 \begin{itemize}
  \item Write your full conditionals out and show that they adhere to the Hammersley-Clifford condition.
 \end{itemize}
  \item \textbf{Metropolis-Hastings}
 \begin{itemize}
  \item Justify your choice of proposal; test different ones if you need to.
 \end{itemize}
\end{itemize}
% 
% \bibliographystyle{apalike}
% \bibliography{refs}

\end{document}          

